Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-02-07 manuscript/manuscript.Rmd
ready to submit version.

Here we used experimental data from a multidisciplinary study of handaxe-making skill acquisition [@pargeter2019; @pargeter2020; @bayani2021] to investigate the interaction between learning processes and the reproduction of specific morphological targets (c.f. "mental template"). This investigation was motivated by the theoretical expectation that, just as developmental processes can act to channel biological variation and shape evolutionary trajectories [@laland2015], learning challenges might influence cultural evolutionary processes [e.g., @schillinger2014]. Results allowed us to identify particular aspects of handaxe morphology that are more and less constrained by learning difficulty, thus helping to partition sources of morphological variation [@lycett2015]. By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach assess the presence and nature of culturally reproduced mental templates. This complements previous work investigated reduction intensity [@shipton2015] and raw material form [@garcía-medrano2019] as alternative explanations for morphological variation and standardization at Boxgrove.

@@

For a decade or so, the mental template concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complimentary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).

Current conceptions of a "mental template" are thus more nuanced than the idea of a fully specified image in the mind of the maker that is directly expressed in material form and transmitted between minds. For example, Hutchence and Scott [-@hutchence2021], leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations. From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

In short, contemporary approaches to the concept of a mental template emphasize the causal importance of production process and constraints and the interaction between individual and group level phenomena. We again note the striking similarity of the perspectives to the concept of "constructive development" as a source of guided variation in evolution biology [@laland2015]. We thus sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.

@@

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

@@

The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The complexity of this issue is further exemplified by the fact that motor skills like knapping cannot be simply learned through observation but must be reconstructed through individual practice using supportive material in social contexts [@stout2017]. The ignorance of this factor becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. As we stated earlier, this reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of knapping skill and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project aims to examine the following research question:

> Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology? To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation. Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model before the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect different level of knapping skill; 3) Throughout the training the novice samples should become more similar to expert samples in both knapping skill and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect. The third assumption here is particularly informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas prepared by Bruce Bradley, which are part of our expert sample as described above. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

@@

Unexpectedly, we found that modern experimental knappers did not closely approximate the PC2 (elongation and pointedness) values of Boxgrove handaxes. More specifically, the Boxgrove assemblage displays an ovate shape featuring a wider tip as previously pointed out by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019], while the experimental assemblages are characterized by a more pointed shape with a longer central axis (**Figure** \@ref(fig:GeneralPCA1)). This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar), but were not provided with a concrete template to copy. It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1) that likely reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182]. Novices then sought to approximate this form, as demonstrated by their instructor and exemplar handaxes from our current expert sample that were given as models, where the mean PC2 value of exemplar handaxes reach 1.18, indicating a high degree of elongation and pointedness even among the expert sample. It should be also noted that the instructor (N. Khreisheh) has learned how to knap and how to teach knapping from one of our expert knappers (B. Bradley), potentially suggesting a cascading effect of social learning that also contributed to a shared mental template between the expert group and the novice group after training. At a higher level, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide a mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise by actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.

Moreover, the pre-training group is unexpectedly similar to the Boxgrove group in PC2 potentially because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of a handaxe and the weight of its corresponding nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to the raw material starting size and/or shape. Achieve handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

@@

Regarding the research question we proposed in the beginning, our case study suggested that the processes of skill learning in a lithic medium do exert biases on the cultural reproduction of artifact morphology. More specifically, skill acquisition has a differential effect on the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so. At a larger theoretical level, it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

Here we used experimental data from a multidisciplinary study of handaxe-making skill acquisition [@pargeter2019; @pargeter2020; @bayani2021] to investigate the interaction between learning processes and the reproduction of specific morphological targets (c.f. "mental template"). This investigation was motivated by the theoretical expectation that, just as developmental processes can act to channel biological variation and shape evolutionary trajectories [@laland2015], learning challenges might influence cultural evolutionary processes [e.g., @schillinger2014]. Results allowed us to identify particular aspects of handaxe morphology that are more and less constrained by learning difficulty, thus helping to partition sources of morphological variation [@lycett2015]. By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach to assess the presence and nature of culturally reproduced mental templates. This complements previous work investigated reduction intensity [@shipton2015] and raw material form [@garcía-medrano2019] as alternative explanations for morphological variation and standardization at Boxgrove.

@@

For a decade or so, the mental template concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016]. A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018]. The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complementary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).

Current conceptions of a "mental template" are thus more nuanced than the idea of a fully specified image in the mind of the maker that is directly expressed in material form and transmitted between minds. For example, Hutchence and Scott [-@hutchence2021], leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations. From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm. However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has value in the study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

In short, contemporary approaches to the concept of a mental template emphasize the causal importance of production process and constraints and the interaction between individual and group level phenomena. We again note the striking similarity of the perspectives to the concept of "constructive development" as a source of guided variation in evolution biology [@laland2015]. We sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.

@@

Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66]. At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners. It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].

When contextual information is less readily available as in Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In addition to measurements that can be applied in almost any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies. These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

@@

The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. Knapping skill and learning difficulties particular to the lithic medium have been less commonly considered as a potential influence of the reprodution of the form. The complexity of this issue is further exemplified by the fact that motor skills like knapping cannot be simply learned through observation but must be reconstructed through individual practice using supportive material in social contexts [@stout2017]. The ignorance of this factor becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. As we stated earlier, this reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of knapping skill and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, our theory-driven data-informed project aims to examine the following research question: Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology? To address this question, we assessed the degree to which trainees succeeded in approximating different aspects of handaxe morphology represented in a sample of modern experts and then compared both samples with archaeological handaxes from Boxgrove.

@@

In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas prepared by Bruce Bradley, which are part of our expert sample as described above. Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013], herself trained by Bruce Bradley, with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage. She was present at all training sessions to provide help and instruction to participants. All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided. All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes. Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping. The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).

@@

Unexpectedly, we found that modern experimental knappers did not closely approximate the PC2 (elongation and pointedness) values of Boxgrove handaxes. More specifically, the Boxgrove assemblage displays an ovate shape featuring a wider tip as previously pointed out by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019], while the experimental assemblages are characterized by a more pointed shape with a longer central axis (**Figure** \@ref(fig:GeneralPCA1)). This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar) but were not provided with a concrete template to copy. It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1). This likely also reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182]. Novices sought to approximate this form, as demonstrated by their instructor and exemplar handaxes. In fact, the four examples presented to trainees have a mean PC2 value of 1.18, indicating a high degree of elongation and pointedness. It should be also noted that the instructor (N. Khreisheh) has learned how to knap and how to teach knapping from one of our expert knappers (B. Bradley), potentially suggesting a cascading effect of social learning that also contributed to a shared mental template between the expert group and the novice group after training. At a higher level, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide a mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise by actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.

The pre-training group is unexpectedly similar to the Boxgrove group in PC2. This is potentially because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113]. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of a handaxe and the weight of its corresponding nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)). It might be worth noting that the expert group is highly variable probably due to the raw material starting size and/or shape. Achieving handaxe forms while removing as little mass as possible (i.e. making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49]. On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

The negative correlation between the PC1 and PC2 values suggests a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form. Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes. In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [-@callahan1979: 90]. As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182]. It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove. Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers. This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms. In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.

@@

Our case study suggested that the processes of skill learning in a lithic medium does in fact exert biases on the cultural reproduction of artifact morphology. More specifically, skill acquisition has a differential effect on the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so. At a larger theoretical level, these results question the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2023-02-07 manuscript/manuscript_word.Rmd
ready to submit version.

---
title: 'Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights'
author:
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY,  USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology. More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill. It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
output:
  word_document:
    reference_docx: "manuscript word.docx"
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

# Introduction

>>>>

---
title: 'Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights'
author:
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY,  USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapping skill and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present new results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differential effect in the cultural reproduction of different aspects of handaxe morphology. More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill. Our findings thus shed new light on how the processes of skill learning can bias the cultural reproduction of artifact morphology.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
output:
  word_document:
    reference_docx: "manuscript word.docx"
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---
# Introduction
